---
title: "Lab #4: Visualization Aesthetics (30 pts)"
author: "Camilla Rosour"
format:
  html:
    code-fold: false
execute: 
  echo: true
knitr:
  opts_chunk: 
    message: false
editor: source
---

## Introduction

Use the code below to load and wrangle the data:

```{r}
#| warning: false
#| document: show
library(tidyverse)
library(RColorBrewer)
library(ggplot2)

hmda_ms_2024 <- read.csv("https://ffiec.cfpb.gov/v2/data-browser-api/view/csv?states=MS&years=2024") |>
  filter(loan_type == 1 & 
           lien_status == 1 & 
           conforming_loan_limit == 'C' & 
           derived_dwelling_category %in% c("Single Family (1-4 Units):Site-Built", 
                                            "Single Family (1-4 Units):Manufactured") & 
           business_or_commercial_purpose == 2 &
           loan_purpose %in% c(1, 2)) |>
  select(lei,
         county_code,
         derived_ethnicity,
         derived_race,
         derived_sex,
         action_taken,
         loan_purpose,
         loan_amount,
         interest_rate,
         property_value,
         loan_to_value_ratio,
         denial_reason.1,
         denial_reason.2,
         denial_reason.3,
         denial_reason.4,
         income,
         debt_to_income_ratio) |>
  mutate(loan_purpose = recode(loan_purpose, 
                               "1" = "Home Purchase",
                               "2" = "Home improvement",
                               .default = NA_character_)) |>
  mutate(action_taken = recode(action_taken, 
                               "1" = "Loan originated", 
                               "2" = "Application approved but not accepted",
                               "3" = "Application denied",
                               "4" = "Application withdrawn by applicant",
                               "5" = "File closed for incompleteness",
                               "6" = "Purchased loan",
                               "7" = "Preapproval request denied",
                               "8" = "Preapproval request approved but not accepted",
                               .default = NA_character_)) |>
  mutate(denial_reason.1 = recode(denial_reason.1, 
                                  "1" = "Debt-to-income ratio",
                                  "2" = "Employment history",
                                  "3" = "Credit history",
                                  "4" = "Collateral",
                                  "5" = "Insufficient cash (downpayment, closing costs)",
                                  "6" = "Unverifiable information",
                                  "7" = "Credit application incomplete",
                                  "8" = "Mortgage insurance denied",
                                  "9" = "Other",
                                  "10" = "Not applicable",
                               .default = NA_character_)) |>
  rowwise() |>
  mutate(denial_reason = case_when(any(!is.na(c_across(c(denial_reason.2,
                                                         denial_reason.3,
                                                         denial_reason.4)))) ~ "Multiple",
      TRUE ~ denial_reason.1)) |>
  ungroup() |>
  mutate(loan_to_value_ratio = as.numeric(loan_to_value_ratio),
         property_value = as.numeric(property_value)) |>
  mutate(debt_to_income_ratio = case_when(debt_to_income_ratio %in% c("36", "37", "38","39","40","41") ~ "36%-41%",
                                          debt_to_income_ratio %in% c("42","43","44","45","46","47","48","49") ~ "42%-49%",
                                          TRUE ~ debt_to_income_ratio)) |>
  mutate(debt_to_income_ratio = factor(debt_to_income_ratio, levels = c("Exempt",
                                                                        "<20%",
                                                                        "20%-<30%",
                                                                        "30%-<36%",
                                                                        "36%-41%",
                                                                        "42%-49%",
                                                                        "50%-60%",
                                                                        ">60%"))) |>
  select(-c(denial_reason.1, 
            denial_reason.2, 
            denial_reason.3, 
            denial_reason.4))

```


::: {.callout-important}
###### Question 1

In the code below, create a bar plot that visualizes how many home loans were for home purchases vs. the number for home improvements. Determine the variable that should appear on the x-axis, along with the geom function that can be used to create a bar plot. Fill in the blanks to ensure your plot has the appropriate context.

```{r}
#| eval: false
hmda_ms_2024 |>
  ggplot(aes(x = loan_purpose)) +
  geom_bar() +
  labs(title = "Home Mortgage Disclosure Act Applications,
Mississippi, 2024",
       x = "Purpose of Home Loan",
       y = "Count of Home Loans") +
  theme_minimal()
##TA Feedback: This graph looks good except for one small labeling mistake the y-variable is count of applications not count of granted loans. -0.5 for small labeling mistake.
##Total for Q1: 2.5 pts
```
:::

::: {.callout-important}
###### Question 2

Create a plot below that counts the number of applicants of each race. Be sure to give it a descriptive title and labels covering all 5 essential components of data context. You can use the code you wrote above as a template to get you started. You'll probably notice that the labels of your plot overlap. You can add `+ coord_flip()` to the end of your code to flip the plot so that the bars are horizontal rather than vertical. This should help with legibility.

```{r}
#| eval: false

hmda_ms_2024 |>
  ggplot(aes(x = derived_race)) +
  geom_bar() +
  labs(title = "Home Mortgage Disclosure Act Applications
by Race, Mississippi, 2024",
       x = "Race",
       y = "Frequency") +
  theme_minimal() +
  coord_flip()
##TA Feedback: This graph looks great except for the same small labeling mistake as before. -0.5 for misslabeling. 
##Total for Q1-Q2: 4 pts
```

:::

::: {.callout-important}
###### Question 3

In the code below, create a histogram that visualizes the distribution of loan amounts. Determine the variable that should appear on the x-axis, along with the geom function that can be used to create a histogram. Set the binwidth to create numeric intervals of 100000. Fill in the blanks to ensure your plot has the appropriate context.

```{r}
#| eval: false
hmda_ms_2024  |> 
  ggplot(aes(x = loan_amount)) +
  geom_histogram(binwidth = 100000) +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississipi, 
2024",
       x = "Loan Amounts Covered/Applied For (USD)",
       y = "Count of Home Loans",
       caption = str_wrap("This histogram displays the distribution of home loan amounts requested by Mississippi applicants in 2024. The plot shows that the majority of applicants requested loans under $250,000. Since the average home loan size in the U.S. in 2024 was $252,505, this indicates that average requested home loan sizes in MS were smaller in comparison to the country as a whole.", 120)) +
  theme_minimal()
##TA Feedback: This graph looks fantastic except for the same labeling mistake as before. -0.5 for labeling mistake.
##Total for Q1-3: 6 pts
```

:::

::: {.callout-important}
###### Question 4

Create a plot below that shows the distribution of loan to value ratios (in intervals of 10). Be sure to give it a descriptive title and labels covering all 5 essential components of data context. Also add a caption to describe the percentage of a loan's property that borrowers most often request when applying for a loan.

You can use the code you wrote above as a template to get you started. Note that there were 20 outlier loans where borrowers requested more than double property's value. There was even one loan where the applicant requested 14 times the property's value. All of these loans were denied. Below I've given you some code to filter out these outliers in order to make the plot more legible. 

```{r}
#| eval: false

hmda_ms_2024 |>
  filter(loan_to_value_ratio < 200) |>
  ggplot(aes(x = loan_to_value_ratio)) +
  geom_histogram(binwidth = 10) +
  labs(title = "Home Mortgage Disclosure Act Applications Loan to Value Ratios,
Mississipi, 2024",
       x = "Loan to Value Ratio",
       y = "Count of Home Loans",
       caption = str_wrap("This histogram displays the distribution of home loan to value rations for requests by Mississippi applicants in 2024. The plot shows that the majority of home loan to value rations lie at between 75 and 100 percent. This indicates that average home loan to value ratios in MS were often close to the full price of the owner's home.", 120)) +
  theme_minimal()
##TA Feedback: This graph looks phenomenal except for the same small labeling mistake as before. -0.5 for minor misslabeling.
##Total for Q1-Q4: 8 pts
```


:::

::: {.callout-important}
###### Question 5

Create a stacked barplot below that counts both the race of the applicants and the action taken in regards to the loan. Be sure to give it a descriptive title and labels covering all 5 essential components of data context. I've given you some code below to move the legend to the bottom of the plot, to reduce the legend size, and to set the number of rows that appear in the legend.

```{r}
#| eval: false

hmda_ms_2024 |>
ggplot(aes(x = derived_race, fill = action_taken)) +
geom_bar() +
  coord_flip() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 6), 
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.4, "cm")) +
  guides(fill=guide_legend(nrow=4)) +
  labs(title = "Stacked Bar Plot of Home Mortgage Disclosure Act Applications
by Race & Action Taken, Mississipi, 2024", y = "Frequency", x = "Race", fill = "Action Taken")
##TA Feedback: This graph look tremendous except for the same labeling mistake. -0.5 for same labeling mistake.
##Total for Q1-Q5: 10 pts
```


::: {.callout-important}
###### Question 6

Copy the plot that you created above into the code chunk below, and then adjust the *position* of the bars to "fill" in order to normalize their height. Add a caption that summarizes your findings. 

```{r}
#| eval: false

hmda_ms_2024 |>
ggplot(aes(x = derived_race, fill = action_taken)) +
geom_bar(position = "fill") +
  coord_flip() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 6), 
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.4, "cm")) +
  guides(fill=guide_legend(nrow=4)) +
  labs(title = "Stacked Bar Plot of Home Mortgage Disclosure Act Applications
by Race & Action Taken, Mississipi, 2024",
y = "Frequency", x = "Race",
fill = "Action Taken",
caption = str_wrap("This stacked bar plot displays the distribution of home loan applications by race, and the action taken in response for requests by Mississippi applicants in 2024. The plot shows that the majority of home loan applications for white and asian applicants were originated, indicating that these racial groups were more likely to have their loans processes.", 120))
##TA Feedback: This graph looks wondrous except for the same labeling mistake. -0.5 for same labeling mistake.
##Total for Q1-6: 12 pts
```



::: {.callout-important}
###### Question 7

Fill in the code below to create a boxplot that visualizes how the distributions of requested loan amounts differed across race. Add a caption that summarizes your findings.  

```{r}
#| eval: false

hmda_ms_2024 |>
  ggplot(aes(x = derived_race, y = loan_amount)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Race of Applicant",
       y = "Loan Amount",
       caption = str_wrap("This stacked box plot displays the distribution of home loan applications by race and loan amounts for requests by Mississippi applicants in 2024. The plot shows that the median value of loan amounts for each group were less than $250,000, but the group that identified as asian had the highest median loan amount. The median loan amount was lowest for those that identified as black or African American, and this group also had the most outliers.", 120)) +
  theme_minimal() 
##TA Feedback: This graph looks marvelous! +2.5 pts
##Total for Q1-Q7: 14.5 pts
```

:::

::: {.callout-important}
###### Question 8

In the code below, fill in the blanks to set the height of each bar to the percent of individuals in a particular race whose mortgage was denied. Select a geom function to create a column chart. Be sure to fill in the correct labels.

```{r}
hmda_denial_by_race <-
  hmda_ms_2024 |>
  group_by(derived_race) |>
  summarize(total_denied = sum(action_taken == "Application denied"),
            percent_denied = sum(action_taken == "Application denied")/n())

hmda_denial_by_race_sex <-
  hmda_ms_2024 |>
  filter(derived_race != "Free Form Text Only") |> #There were only 3 denials with free form text for race in MS in 2024, and they have been removed here for plot legibility
  group_by(derived_race, derived_sex) |>
  summarize(total_denied = sum(action_taken == "Application denied"),
            percent_denied = sum(action_taken == "Application denied")/n())

hmda_denial_by_race
```

```{r}
#| eval: false

hmda_denial_by_race |>
  ggplot(aes(x = derived_race, y = percent_denied)) +
  geom_col() +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Race of Applicant",
       y = "Percent Denied",
       caption = str_wrap("This column plot visualizes the denial rates of home mortgage applications in MS in 2024. The plot shows that Black or African American applicants, American Indian or Alaska Native applicants, or applicants of 2 or more minority races were all denied mortgage at double the rate of White applicants. This may indicate disparities in lending practices.", 120)) +
  theme_minimal() +
  coord_flip()
##TA Feedback: This graph looks wondrous! +2.5 
##Total for Q1-8: 17 pts
```

:::


::: {.callout-important}
###### Question 9

Copy your code above, replacing the `hmda_denial_by_race` data frame with `hmda_denial_by_race_sex`. Using the `fill` aesthetic, add the variable for describing the applicants' sex to the plot. Set the position of the columns to `"dodge"`, and be sure to add a label for the `fill` aesthetic. Adjust the caption to summarize your findings.

```{r}
#| eval: false

hmda_denial_by_race_sex |>
  ggplot(aes(x = derived_race, y = percent_denied, fill = derived_sex)) +
  geom_col() +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Race of Applicant",
       y = "Percent Denied",
       fill = "Sex",
       caption = str_wrap("This column plot visualizes the denial rates of home mortgage applications in MS in 2024 by sex and race. The plot shows that for most racial groups, females were denied more than males, and joint applications did better on average. The data is harder to generalize because of the missing data on sex, but it seems that female's are at a disadvantage.", 120)) +
  theme_minimal() +
  coord_flip()
##TA Feedback: This graph looks terrific! +2.5
##Total for Q1-9: 19.5 pts
```

:::

::: {.callout-important}
###### Question 10

Create a plot below that visualizes how many applicants *of each race* were denied mortgages *for each denial reason*. Be sure to give it a descriptive title and labels covering all 5 essential components of data context. You may wish to borrow some of your previous code to adjust the plot for legibility. I've given you some code below to filter the data frame to just those mortgage applications that were denied.

```{r}
#| eval: false

hmda_ms_2024 |>
  filter(action_taken == "Application denied") |>
#Fill remaining lines here
  ggplot(aes(x = denial_reason)) +
  geom_bar() +
  labs(title = "Home Mortgage Disclosure Act Applications Denial Reasons
by Race, Mississippi, 2024",
       x = "Denial Reason",
       y = "Frequency",
       fill = "Race of Applicant") +
  theme_minimal() +
  coord_flip() +
  facet_wrap(vars(derived_race))
##TA Feedback: This graph looks extraordinary! +2.5 
##Total for Q1-10: 22 pts
```

:::

::: {.callout-important}
###### Question 11

Below I've given you some code to produce a plot that visualizes the percent of mortgage applications that resulted in each action (e.g. loan originated, denied, etc.), broken out by the debt-to-income ratio of the applicant. Referring to last week's lab, add a line to this code to facet this plot by race. For which races was the denial rate higher for applicants with a lower than 50% debt-to-income ratio? Adjust the fact and interpretation presented in the caption with your response.

```{r}
#| eval: true


# The denial rate was higher for applicants with a lower than 50% debt-to-income ratio for black/afican american applicants and american indian or alaska native applicants. 

hmda_ms_2024 |>
  ggplot(aes(x = debt_to_income_ratio, fill = action_taken)) +
  geom_bar(position = "fill") +
  coord_flip() +
  facet_wrap(vars(derived_race)) +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Debt-to-Income Ratio",
       y = "Percent of Mortgage Applications",
       fill = "Action Taken", 
       caption = str_wrap("This bar plot visualizes the percent of mortgage applications that resulted in each action, broken out by the debt-to-income ratio of the applicant. From this plot, we can see that most applicants with a debt-to-income ratio of 50% or lower were denied loans at higher rates if they identified as black/African American or American Indian or Alaska Native. This shows that banks are less likely to lend to applicants of these racial categories even when their debt-to-income ratio is less than 50%.",width = 120)) +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8), 
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.4, "cm"),
        plot.caption = element_text(size = 8)) +
  guides(fill=guide_legend(nrow = 3))
##TA feedback: This graph looks beautiful! +2.5
##Total for Q1-Q11: 24.5 pts
```

:::


::: {.callout-important}
###### Question 12

Below I've given you some code to produce a plot that visualizes how the distributions of interest rates differed across debt-to-income ratios when loans were approved. Referring to last week's lab, add a line to this code to facet this plot by race. For which races were the interest rates higher even with lower debt-to-income ratios?  Adjust the fact and interpretation presented in the caption with your response.


```{r}
#| eval: true

# interest rates were higher for black/african-american applicants and those with 2 or more minority races even with lower debt=to=income-ratios. fot those that are white with lower debt-to-income ratios, there were many higher interest rates that were outliers.

hmda_ms_2024 |>
  filter(action_taken == "Loan originated") |>
  ggplot(aes(x = debt_to_income_ratio, y = as.numeric(interest_rate))) +
  geom_boxplot() +
  coord_flip() +
  facet_wrap(vars(derived_race)) +
  labs(title = "Home Mortgage Disclosure Act Applications, Mississippi, 2024",
       x = "Debt-to-Income Ratio",
       y = "Interest Rate",
       caption = str_wrap("These boxplots visualize the distribution of interest rates for loans originated in MS in 2024, broken out by the debt-to-income ratio of the applicant. From this plot, we can see that the median interest rate for borrowers with a debt-to-income ratio of under 50% was around 7%. Interest rates were slightly higher for those that identified as black or African American, even with lower debt-to-income ratios. This shows that though a high debt-to-income ratio may be one factor contributing to higher interest rates, race might also be a factor.",width = 120)) +
  theme_minimal()
##TA Feedback: This graph looks awesome! +2.5
##Total for Q1-12: 27 pts
  
```

:::

::: callout-caution
## Ethical Considerations

This dataset makes loan-level data about individual sensitive financial information available to the public. To reduce the risk that borrowers may be identified in the data, the CFPB removes certain variables before publication, such as the loan's unique identifier, the property address, the applicant's credit score, and any free-form text that the applicant writes on the application. Applicants do not have the option to opt-out of their data being disclosed in this dataset. If it were possible to re-identify individual applicants in this dataset, what would be the risks to the public disclosure of their information? Which applicants would be the most vulnerable? How should we think about the balance between the social value of this public data disclosure and the possible risks to those represented in it?  Share your ideas on our `discussions` Slack channel.
##TA Grading: Final Score: 27/30 or 90%
:::
